Exploring the Role of Small Molecules in Biological Systems Using Network Approaches

### 2015-03-22 paper.tex
updated with increased emphasis on R packages for network analysis

Using these databases, network methods can be applied at various
stages of the drug discovery pipeline. For example, target
identification can benefit from network analyses by characterizing the
role of a given target in the context of the larger interaction and
regulatory network. Thus targeting proteins that have a high degree of
connectedness (as measured by essentiality and centrality) may lead to
significant side effects in other parts of the system and can be the
source of observed side effects \cite{Wang:2013gn}. On these lines,
approaches to characterize toxicity have employed network analyses
\cite{Kim:2014xs,Zhang:2014hj}. While some approaches have considered
side effects [REF] others have considered the problem at the genomic
level by characterizing gene networks that are perturbed by small
molecules. Campaigns focused on combination therapies can also benefit
from a network representation and network inference approaches
\cite{Huang:2014by,Tang:2013lr} to identify target combinations whose
simultaneous inhibition leads to synergistic effects. Mechanism of
action studies can benefit from network analyses by integrating
multiple data types. An example of this was described by Young et al
\cite{Young:2008aa} that integrated high content screening and
ligand-target prediction models to construct network representations
of MoA hypotheses. Network approaches have also been proposed for SAR
analysis \cite{Iyer:2011ij,Wawer:2008aa}, usually at the lead
optimization stage. Drug repositioning efforts
\cite{Dudley:2011dn,Gottlieb:2011rx} benefit from network analyses
such as community detection algorithms \cite{girvan2002community}.

@@

@@

Schadt et al \cite{Schadt:2009zl} have stressed the importance of
viewing diseases as manifestations of perturbed biological networks
and consequently, the importance of drug discovery approaches that
address the networks underlying a disease state.. Network approaches
to exploring diseases have traditionally explored biological networks
(e.g., protein-protein interaction or gene regulatory networks) in the
context of individual diseases \cite{Lim:2006fe,
  Altieri:2008aa}. Other approaches on the other hand have focused on
classes of disease \cite{Zhang:2013tx} or even all known human
diseases \cite{Goh:2007aa} (the so called ``diseaseome''). While most
of these studies have primarily focused on identifying relationships
between diseases in different ways such as sharing implicated genes
\cite{Bauer-Mehren:2011tv,Goh:2007aa}, sharing symptoms
\cite{Zhou:2014qp} or based on bibliometric associations
\cite{Zhang:2014sh}. Studies on disease networks tend to construct
heterogenous (or $k$-partite) networks where nodes can represent
different classes of entities. An example is the network constructed
from long non-coding RNAs and diseases by Zhou et al
\cite{Zhou:2015if}. Based on this network, a random walk approach was
employed to identify novel lncRNA-disease associations. These
approaches play an important role in drug discovery as they provide
opportunities for repurposing drugs based on novel disease-target or
disease-pathway associations. Another view provided by such analysis
is the druggability of a disease; that is, based on underlying targets
and pathways, is it even feasible to treat the disease using a small
molecule intervention \cite{Boran:2010fk}. An important function of
network analyses of diseases is to go beyond simply identifying new
targets for diseases or links between diseases, but instead attempt to
elucidate the mechanisms underlying condition. As an example of this is
the paradoxical effects of MEK inhibition and PI3K/AKT inhibition and
the effects on cancer resistance \cite{Sos:2009gc}. Together with knowledge
of the underlying network connectivity, these approaches allow one to
design combination treatments - multiple drugs that affect multiple
targets and pathways. This approach was exemplified by Lehar et al
\cite{Lehar:2009gu} who proposed that drug combination response
surfaces were functionally related to the underlying connectivity of
the drug targets.

@@

One application of this dataset is to explore the nature of fragments
associated with a document, and for each of those fragments identify
the documents associated with them. This procedure can be repeated to
construct scaffold-document network that represents the chemical and
bibliometric space associated with a starting document. As an example
of this type of analysis we constructed a network starting from a
report describing a series of imidazonaphthalimides and their
antitumor activity \cite{Brana:2002pr}. The network was limited to 3
iterations (document $\rightarrow$ fragment $\rightarrow$ document
$\rightarrow$ fragment). In addition we applied constraints on the
fragments (e.g., must be contained in 20 to 50 molecules and exhibit a
minimum complexity), resulting in a network with 1277 nodes (816
document nodes and 461 fragment nodes) shown in Figure
\ref{fig:scafodcnet}. 
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{img/12477366}
  \caption{An example of a scaffold-document network, limited to three
    iterations. Squares represent fragments and circles represent
    documents. Document nodes are colored based on their major topic
    MeSH heading. The seed document is shown with a red border (top
    left).}
  \label{fig:scafdocnet}
\end{figure}
In this figure, fragment nodes are squares and document nodes are
circles, colored by the major topic MeSH heading. (Only the 12 most
frequent MeSH headings are considered - document nodes outside this
set are colored grey). Though difficult to navigate as a static  
figure, the network structure is interesting in that the immediate
neighborhood of the seed document (red border, top left) is relatively
small. The median number of fragments associated with a document is 1,
(though two documents were connected to more than 100
fragments). Conversely, the median number of documents connected to a
fragment was 3. Fragments tend to be associated with multiple
documents (which is expected based on the number of iterations used to
construct the network, as well as the constraints on fragment
characteristics) and based on the MeSH headings there is topical
overlap between many of these documents. For example, most document
clusters are annotated with the \emph{Chemistry} and \emph{Chemical
  synthesis} MeSH headings. Given that the seed document has been
annotated with \emph{Chemical synthesis} it is not surprising that
based on the chemistry contained within it, related documents will
also be related to synthesis. Importantly, if one considers all
documents in ChEMBL (with valid Pubmed ID's), the two most frequent
MeSH headings are \emph{Chemistry} and \emph{Chemical synthesis} and
thus overrepresentation of these terms is unsurprising.

@@

The core concepts involved in computational drug discovery can be
reduced to genes, proteins, pathways and small molecules interacting
with them. Computational representations of these concepts and the
methods to be manipulate them have been implemented in a variety of
software tools. A comprehensive computational drug discovery platform
will include a number of these tools, as it is rare that a single tool
addresses all aspects.

In this sense, the R environment presents a robust platform for
computational drug discovery, especially the stages up to and
including lead optimization. The Bioconductor set of packages provide
a plethora of tools to analyze a variety of ``omics'' data. Examples
include packages that support gene expression (microarray or RNAseq)
analyses, tools for differential expression analysis, gene and pathway
enrichment analysis and so on.

In contrast, support for small molecule data is limited. Currently,
there are two packages that focus on representing and manipulating
chemical structures - \texttt{ChemmineR} \cite{Cao:2008fj} and \rcdk
\cite{Guha:2007aa}. Both expose cheminformatics functionality
implemented in an underlying Java library (JOELib for ChemmineR and
the CDK for \rcdk). In this section we focus on the use of \rcdk as a
means to manipulate and analyze chemical structure data.

@@

mols <- parse.smiles(mipe$SMILES_STD)
fps <- lapply(mols, get.fingerprint, type='extended')
fpsim <- fp.sim.matrix(fps)
colnames(fpsim) <- mipe$SAMPLE_NAME
rownames(fpsim) <- mipe$SAMPLE_NAME
fpsim[ fpsim < 0.6 ] <- 0
\end{lstlisting}
In this code snippet, we compute a pairwise similarity matrix that we
will eventually use as an adjacency matrix to construct the similarity
network. We only keep those similarity values greater than 0.6 so as
to focus only on similar molecules.

>>>>

Using these databases, network methods can be applied at various
stages of the drug discovery pipeline. For example, target
identification can benefit from network analyses by characterizing the
role of a given target in the context of the larger interaction and
regulatory network. Thus targeting proteins that have a high degree of
connectedness (as measured by essentiality and centrality) may lead to
significant side effects in other parts of the system and can be the
source of observed side effects \cite{Wang:2013gn}. On these lines,
approaches to characterize toxicity have employed network analyses
\cite{Kim:2014xs,Zhang:2014hj}. While some approaches have considered
side effects \cite{Lounkine:2012yf} others have considered the problem
at the genomic level by characterizing gene networks that are
perturbed by small molecules. Campaigns focused on combination
therapies can also benefit from a network representation and network
inference approaches \cite{Huang:2014by,Tang:2013lr} to identify
target combinations whose simultaneous inhibition leads to synergistic
effects. Mechanism of action studies can benefit from network analyses
by integrating multiple data types. An example of this was described
by Young et al \cite{Young:2008aa} that integrated high content
screening and ligand-target prediction models to construct network
representations of MoA hypotheses. Network approaches have also been
proposed for SAR analysis \cite{Iyer:2011ij,Wawer:2008aa}, usually at
the lead optimization stage. Drug repositioning efforts
\cite{Dudley:2011dn,Gottlieb:2011rx} benefit from network analyses
such as community detection algorithms \cite{girvan2002community}.

@@

\section{R for Network Analyses}
\label{sec:r-network-analyses}

While this chapter focuses on the use of network methodologies in
various drug discovery related problems, we also focus on the use of R
for these analyses. Section \ref{sec:r-as-platform} discusses R
packages oriented towards handling molecular data types such as
chemical structures and genomic data. In this section we briefly
review the packages that are available for handling and analyzing
network datasets. The \texttt{igraph} \cite{Csardi:2006lp} and
\texttt{network} \cite{Butts:2008tp} packages represent comprehensive
tools to construct, manipulate and visualize network datasets within
the R environment. The \texttt{graph} package \cite{Gentleman:1wt}
also aims to support general graph manipulation, but is somewhat
limited in scope compared to \texttt{igraph} and \texttt{network}
(both of which share some aspects of syntax and interfaces). These
three packages are the basis of many other network analysis and visualization
packages. An example is \texttt{statnet} \cite{Handcock:2008il} which
employs the \texttt{network} package and focuses on the use of
exponential random graph models \cite{Robins:2007cr} to support
statistical inference and simulation.

While many network analysis packages in R focus on social science
applications, there are a plethora of R packages that focus on
biological networks. For example, the WCGNA package
\cite{Langfelder:2008cj} implements methods to construct gene networks
based on co-expression data. The Corbi package \cite{Huang:2013sy}
implements the CNetQ algorithm for network alignment, a task that is
relevant for predicting new protein-protein interactions. Given the
integrative nature of many biological problems that require the use of
multiple data types (e.g., genomic, proteomic, small molecule),
network approaches are commonly applied to investigating them. A
number of R packages support such integrative analyses - BioNet
\cite{Beisser:2010xp} is an R package that supports a variety of
integrative analyses, with a particular focus on identifying
functional modules in $k$-partite networks. NetWeAvers
\cite{McClellan:2013zf} is similar in nature, though it has
specialized support for mass spectrometry data. A number of packages
support statistical operations on graphs. For example,
\texttt{DEGraph} implements a method to test for differential
expression of gene networks \cite{Jacob:2012yg}. The \texttt{SANTA}
package \cite{Cornish:2014vs} implements methods to test the
association between a network and a phenotype. Several packages are
also available for enrichment analyses including \texttt{FGNet},
\texttt{miRNApath} and \texttt{viper}.

@@

Schadt et al \cite{Schadt:2009zl} have stressed the importance of
viewing diseases as manifestations of perturbed biological networks
and consequently, the importance of drug discovery approaches that
address the networks underlying a disease state.. Network approaches
to exploring diseases have traditionally explored biological networks
(e.g., protein-protein interaction or gene regulatory networks) in the
context of individual diseases \cite{Lim:2006fe,
  Altieri:2008aa}. Other approaches on the other hand have focused on
classes of disease \cite{Zhang:2013tx} or even all known human
diseases \cite{Goh:2007aa} (the so called ``diseaseome''). While most
of these studies have primarily focused on identifying relationships
between diseases in different ways such as sharing implicated genes
\cite{Bauer-Mehren:2011tv,Goh:2007aa}, sharing symptoms
\cite{Zhou:2014qp} or based on bibliometric associations
\cite{Zhang:2014sh}. Studies on disease networks tend to construct
heterogenous (or $k$-partite) networks where nodes can represent
different classes of entities. An example is the network constructed
from long non-coding RNAs and diseases by Zhou et al
\cite{Zhou:2015if}. Based on this network, a random walk approach was
employed to identify novel lncRNA-disease associations. These
approaches play an important role in drug discovery as they provide
opportunities for repurposing drugs based on novel disease-target or
disease-pathway associations. For example, Pacini et al
\cite{Pacini:2013kl} describe DvD, a pipeline comprised of R and
Cytoscape \cite{Shannon:2003aa} to identify repurposing opportunities
using gene expression profiles. Another view provided by such analysis
is the druggability of a disease; that is, based on underlying targets
and pathways, is it even feasible to treat the disease using a small
molecule intervention \cite{Boran:2010fk}. An important function of
network analyses of diseases is to go beyond simply identifying new
targets for diseases or links between diseases, but instead attempt to
elucidate the mechanisms underlying condition. As an example of this
is the paradoxical effects of MEK inhibition and PI3K/AKT inhibition
and the effects on cancer resistance \cite{Sos:2009gc}. Together with
knowledge of the underlying network connectivity, these approaches
allow one to design combination treatments - multiple drugs that
affect multiple targets and pathways. This approach was exemplified by
Lehar et al \cite{Lehar:2009gu} who proposed that drug combination
response surfaces were functionally related to the underlying
connectivity of the drug targets.

@@

One application of this dataset is to explore the nature of fragments
associated with a document, and for each of those fragments identify
the documents associated with them. This procedure can be repeated to
construct scaffold-document network that represents the chemical and
bibliometric space associated with a starting document. As an example
of this type of analysis we constructed a network starting from a
report describing a series of imidazonaphthalimides and their
antitumor activity \cite{Brana:2002pr}. The network was limited to 3
iterations (document $\rightarrow$ fragment $\rightarrow$ document
$\rightarrow$ fragment). In addition we applied constraints on the
fragments (e.g., must be contained in 20 to 50 molecules and exhibit a
minimum complexity), resulting in a network with 1277 nodes (816
document nodes and 461 fragment nodes) shown in Figure
\ref{fig:scafdocnet}.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{img/12477366}
  \caption{An example of a scaffold-document network, limited to three
    iterations. Squares represent fragments and circles represent
    documents. Document nodes are colored based on their major topic
    MeSH heading. The seed document is shown with a red border (top
    left).}
  \label{fig:scafdocnet}
\end{figure}
In this figure, fragment nodes are squares and document nodes are
circles, colored by the major topic MeSH heading. (Only the 12 most
frequent MeSH headings are considered - document nodes outside this
set are colored grey). Though difficult to navigate as a static  
figure, the network structure is interesting in that the immediate
neighborhood of the seed document (red border, top left) is relatively
small. The median number of fragments associated with a document is 1,
(though two documents were connected to more than 100
fragments). Conversely, the median number of documents connected to a
fragment was 3. Fragments tend to be associated with multiple
documents (which is expected based on the number of iterations used to
construct the network, as well as the constraints on fragment
characteristics) and based on the MeSH headings there is topical
overlap between many of these documents. For example, most document
clusters are annotated with the \emph{Chemistry} and \emph{Chemical
  synthesis} MeSH headings. Given that the seed document has been
annotated with \emph{Chemical synthesis} it is not surprising that
based on the chemistry contained within it, related documents will
also be related to synthesis. Importantly, if one considers all
documents in ChEMBL (with valid Pubmed ID's), the two most frequent
MeSH headings are \emph{Chemistry} and \emph{Chemical synthesis} and
thus overrepresentation of these terms is unsurprising.

@@

The core concepts involved in computational drug discovery can be
reduced to genes, proteins, pathways and small molecules interacting
with them. Computational representations of these concepts and the
methods to be manipulate them have been implemented in a variety of
software tools. These range from software libraries such as the CDK
\cite{Steinbeck:2006aa} or RDKit \cite{RDKit:2008aa} for manipulating
chemical structures to pipeline tools such as KNIME
\cite{KNIME:2008aa} or PipelinePilot that can be used to integrate
disparate tools and data sources using a graphical user interface.  A
comprehensive computational drug discovery infrastructure will include
a number of these tools, as it is rare that a single tool addresses
all aspects.

Statistical modeling and machine learning methods have become an
integral part of computational drug discovery. Many drug discovery
tools and platforms provide access to statistical methods. In some
cases such as in RDKit, these are implemented from scratch, where as
in others such as PipelinePilot, external statistical tools such as R
and JMP are integrated.

If one is primarily interested in the statistical analyses in a drug
discovery setting, the R environment presents a robust platform for
computational drug discovery, especially the stages up to and
including lead optimization.  It is well known that many cutting edge
statistical techniques are first implemented in R and this is
increasingly becoming the case for much of bioinformatics. The
Bioconductor set of packages provides a plethora of tools to analyze a
variety of ``omics'' data. Examples include packages that support gene
expression (microarray or RNAseq) analyses, tools for differential
expression analysis, gene and pathway enrichment analysis and so on.

However, support for small molecule data is limited. A number of
packages focus on specific cheminformatics related
tasks. Examples include the \texttt{Rcpi} package that computes
protein ligand descriptors \cite{Cao:2015uq} , the \texttt{QuACN} package that
computes a number of graph invariants \cite{Mueller:2011rz} from a
molecular graph and the \texttt{RMol} package to parse MOL files
\cite{Grabner:2012rr}. However, in many scenarios it is useful to
be able to manipulate chemical structures at a lower level. Thus one
requires access to core cheminformatics functionality. Currently,
there are two packages that focus on representing and manipulating
chemical structures - \texttt{ChemmineR} \cite{Cao:2008fj} and \rcdk
\cite{Guha:2007aa}. Both expose cheminformatics functionality
implemented in an underlying Java library (JOELib for ChemmineR and
the CDK for \rcdk). These packages provide a low level interface to
cheminformatics libraries, allowing the R user to manipulate data
structures representing atoms, bonds and molecules as well as perform
core cheminformatics tasks such as atom typing and aromaticity
perception. But in addition, higher level features are also made
available including descriptor calculation, substructure searching,
fragmentation and 2D structure depictions. In this section we focus on
the use of \rcdk a means to manipulate and analyze chemical structure
data.

@@

mols <- parse.smiles(mipe$SMILES_STD)
fps <- lapply(mols, get.fingerprint, type='extended')
fpsim <- fp.sim.matrix(fps)
colnames(fpsim) <- mipe$SAMPLE_NAME
rownames(fpsim) <- mipe$SAMPLE_NAME
fpsim[ fpsim < 0.6 ] <- 0
\end{lstlisting}
In this snippet, we computed hashed path fingerprints as implemented
in the CDK, using the default values of length (1024) and depth
(6). The CDK supports multiple fingerprint types including
substructural keys (MACCS, Pubchem) and circular fingerprints (ECFP)
and we can easily switch to another one by changing the \texttt{type}
argument. Once the fingerprints are computed we compute a pairwise
similarity matrix that we will eventually use as an adjacency matrix
to construct the similarity network. We only keep those similarity
values greater than 0.6 so as to focus only on similar molecules.
